### 2016-02-08 prepReport/theorBack.tex
Structured final document

\chapter{Theoretical Background}\label{chap:theorBack}

\section{Basic Concepts of Machine Learning}
Machine Learning is a field of Computer Science that studies the development of mathematical techniques that allow software to learn autonomously, without an explicit description of each rule of operation. Its goal is to extract latent features from the data that allow an immediate classification of each input data into a particular class -- the catch is that there is no previous rule formulation, but instead we have an adaptive model that adjusts is parameters accordingly with the input data it receives, improving the estimates it yields as it receives new input samples.

@@

\section{Artificial Neural Networks}\label{sec:theorBack_ann}

@@

\section{Recurrent Neural Networks}\label{sec:theorBack_rnn}
A Recurrent Neural Network (RNN) is, essentially, a regular ANN where some neurons (especially in the hidden layer) have \emph{feedback connections} to themselves, i.e. their outputs are fed as inputs. The relevance of this different structure is the possibility to retain \emph{sequence} information about the data. Before, each incoming data point only contributed to the training of the network, but no information about the correlation between themselves and the data points that preceded them did not influence the training step. They were regarded as if no temporal relationship existed, and therefore each data point is conditionally independent of any other. This is obviously not necessarily truth, and in fact there are many cases, where the correlation between data points is high for those closely spaced in time, in which it is actually completely false, as in video signals, audio signals, or other kinds of \emph{temporal sequences} of data. Therefore, the feedback connection of the neuron to himself acts as a kind of \emph{memory element} that takes into account in the present decision, the history of decisions previously taken, and hence the previous data. 
Figure~\ref{fig:recneuron} suggests a possible structure for a neuron of a hidden layer in an RNN, and also an alternate representation, where the structure is unfold through time.

@@

\section{Long Short-Term Memory Networks}\label{sec:theorBack_lstm}
To overcome the issue of failing to remember long-term dependencies, Hochreiter and Schmidhüber proposed, in 1997, a novel approach to the RNNs called the \textit{Long-Short Term Network}~\cite{Hoch97}. This section explains the main idea of this approach~\ref{sec:struct_lstm}, and also how it is trained\ref{sec:training_lstm}, serving as a support basis for the work of this thesis.
Although originally formulated in 1997, its formulation has been incrementally updated in~\cite{Gers00} and~\cite{Gers2000}, and the most current version is the one in~\cite{Graves05}. One of the inital proposers of LSTM, Prof. Jürgen Schmidhüber, did a survey on the most common variations of the structure~\cite{Greff15} last year, and this will the basis of this short theoretical presenation, as well as the work that will be developed in this thesis.

\subsection{Structure, Operation and Equations}\label{sec:struct_lstm}
A single LSTM neuron is presented in Figure~\ref{fig:lstmneuron}. As we can see from the picture, we still have the recurrent connections from the regular RNNs, but now there are multiple entry points that control the flow of information through the network. Although ommited from the  picture, all the gates are biased, as is suggested in Equations~\ref{eq:equationsLstm}. The main components, their role and relevance, are explained as follows

\begin{itemize}

    \item \textbf{Input Gate} -- this is the input gate, where the importance of each feature of the input vector at time $t$, $\mb{x}^t$, and the output vector at the previous time step $\mb{y}^{t-1}$ is weighed in, producing an output $\mb{i}^{t}$.

@@

After this small conceptual definitions, that allow us to grasp some intuition on the operation of a \textit{single} LSTM cell, I can present the overview of a \textit{layer} of LSTM neurons and their formal mathematical formulation, that will be needed both for the high-level model and the HDL description. The operation of each gate of the layer is given by the following set of equations

\begin{align}
    \mb{z}^{(t)} & = g(\mb{W}_z \mb{x}^{(t)} + \mb{R}_z \mb{y}^{(t-1)} + \mb{b}_z) \\
    \mb{i}^{(t)} & = \sigma(\mb{W}_i \mb{x}^{(t)} + \mb{R}_i \mb{y}^{(t-1)} + \mb{p}_i \odot \mb{c}^{(t-1)} + \mb{b}_i) \\
    \mb{f}^{(t)} & = \sigma(\mb{W}_f \mb{x}^{(t)} + \mb{R}_f \mb{y}^{(t-1)} + \mb{p}_f \odot \mb{c}^{(t-1)} + \mb{b}_f) \\
    \mb{c}^{(t)} & = \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \\
    \mb{o}^{(t)} & = \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \\
    \mb{c}^{(t)} & = \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) 
\end{align}
The $i$-th element of the previous vectors in bold corresponds to the value of the gate of the $i$-th neuron of the layer, which is a very convenient and compact representation of the whole layer. Although useful for a high-level description in a programming language, the matrix representation  may not be suitable for a direct HDL port. In that case, all the above equations still hold, but instead of \textit{vectors} of outputs, we will have a single output, and we only use the appropriate value in the weight matrices $\mb{W}_{\cdot}$ and $\mb{R}_{\cdot}$
\begin{figure}
	\centering
	\includegraphics[width=0.9\linewidth]{figures/lstmneuron.png}
    \caption{A complete LSTM neuron, with all the features as described in~\cite{Graves05}}
	\label{fig:lstmneuron}
\end{figure}

\subsection{Training -- SPSA}\label{sec:training_lstm}
Since this work aims to perform on-chip learning, it is important to find a suitable learning scheme. Since I am aiming at an hardware implementation, and although the memory resources of nowadays FPGAs are abundant, one must find an algorithm that uses the less memory as possible (at the smallest performance cost possible), in order to use the additional memory, for instance, to add additional LSTM cells that can improve the performance of our system. 
That being said, we see that the usual training algorithms for LSTM cells -- i.e. Real-Time Recurrent Learning (RTRL), Truncated Backpropagation Through Time (BPTT) or a mixture of both~\cite{Graves05}~\cite{Hoch97}~\cite{Greff15} -- usually involve the storage of the deltas of each layer for every time instant in the training epoch (from $0$ to $T$), which is a highly non-scalable solution both in terms of memory and performance. A most efficient approach to training times series dependant structures like LSTM is the use of \textbf{Simultaneous Perturbation Stochastic Approximation} (SPSA)~\cite{Spall98}. The weight update for the $i$-th weight at the time step $t$ is given by

>>>>

\chapter{Problem Characterization}\label{chap:theorBack}

\section{Theoretical Background}

\subsection{Basic Concepts of Machine Learning}
Machine Learning is a field of Computer Science that studies the development of mathematical techniques that allow software to learn autonomously, without an explicit description of each rule of operation. Its goal is to extract latent features from the data that allow an immediate classification of each input data into a particular class -- the catch is that there is no previous rule formulation, but instead we have an adaptive model that adjusts is parameters accordingly with the input data it receives, improving the estimates it yields as it receives new input samples.

@@

\subsection{Artificial Neural Networks}\label{sec:theorBack_ann}

@@

\subsection{Recurrent Neural Networks}\label{sec:theorBack_rnn}
A Recurrent Neural Network (RNN) is, essentially, a regular ANN where some neurons (especially in the hidden layer) have \emph{feedback connections} to themselves, i.e. their outputs are fed as inputs. The relevance of this different structure is the possibility to retain \emph{sequence} information about the data. Before, each incoming data point only contributed to the training of the network, but no information about the correlation between themselves and the data points that preceded them did not influence the training step. They were regarded as if no temporal relationship existed, and therefore each data point is conditionally independent of any other. This is obviously not necessarily truth, and in fact there are many cases, where the correlation between data points is high for those closely spaced in time, in which it is actually completely false, as in video signals, audio signals, or other kinds of \emph{temporal sequences} of data. Therefore, the feedback connection of the neuron to himself acts as a kind of \emph{memory element} that takes into account in the present decision, the history of decisions previously taken, and hence the previous data. 
Figure~\ref{fig:recneuron} suggests a possible structure for a neuron of a hidden layer in an RNN, and also an alternate representation, where the structure is unfold through time.

@@

\subsection{Long Short-Term Memory Networks}\label{sec:theorBack_lstm}
To overcome the issue of failing to remember long-term dependencies, Hochreiter and Schmidhüber proposed, in 1997, a novel approach to the RNNs called the \textit{Long-Short Term Network}~\cite{Hoch97}. This section explains the main idea of this approach~\ref{sec:struct_lstm}, and also how it is trained\ref{sec:training_lstm}, serving as a support basis for the work of this thesis.
Although originally formulated in 1997, its formulation has been incrementally updated in~\cite{Gers00} and~\cite{Gers2000}, and the most current version is the one in~\cite{Graves05}. One of the inital proposers of LSTM, Prof. Jürgen Schmidhüber, did a survey on the most common variations of the structure~\cite{Greff15} last year, and this will the basis of this short theoretical presenation, as well as the work that will be developed in this thesis.

\subsubsection{Structure, Operation and Equations}\label{sec:struct_lstm}
A single LSTM neuron is presented in Figure~\ref{fig:lstmneuron}. As we can see from the picture, we still have the recurrent connections from the regular RNNs, but now there are multiple entry points that control the flow of information through the network. Although ommited from the  picture, all the gates are biased, as is suggested in Equations~\ref{eq:equationsLstm}. The main components, their role and relevance, are explained as follows

\begin{figure}[H]
	\centering
	\includegraphics[width=0.9\linewidth]{figures/lstmneuron.png}
    \caption{A complete LSTM neuron, with all the features as described in~\cite{Graves05}}
	\label{fig:lstmneuron}
\end{figure}

\begin{itemize}
    \item \textbf{Input Gate} -- this is the input gate, where the importance of each feature of the input vector at time $t$, $\mb{x}^t$, and the output vector at the previous time step $\mb{y}^{t-1}$ is weighed in, producing an output $\mb{i}^{t}$.

@@

After these small conceptual definitions, that allow us to grasp some intuition on the operation of a \textit{single} LSTM cell, I can present the overview of a \textit{layer} of LSTM neurons and their formal mathematical formulation, that will be needed both for the high-level model and the HDL description. The operation of each set of gates of the layer is given by the following set of equations

\begin{align}\label{eq:equationsLSTM}
    \mb{z}^{(t)} & = g(\mb{W}_z \mb{x}^{(t)} + \mb{R}_z \mb{y}^{(t-1)} + \mb{b}_z) \\
    \mb{i}^{(t)} & = \sigma(\mb{W}_i \mb{x}^{(t)} + \mb{R}_i \mb{y}^{(t-1)} + \mb{p}_i \odot \mb{c}^{(t-1)} + \mb{b}_i) \\
    \mb{f}^{(t)} & = \sigma(\mb{W}_f \mb{x}^{(t)} + \mb{R}_f \mb{y}^{(t-1)} + \mb{p}_f \odot \mb{c}^{(t-1)} + \mb{b}_f) \\
    \mb{c}^{(t)} & = \mb{i}^{(t)} \odot \mb{z}^{(t)} + \mb{f}^{(t)} \odot \mb{c}^{(t-1)} \\
    \mb{o}^{(t)} & = \sigma(\mb{W}_o \mb{x}^{(t)} + \mb{R}_o \mb{y}^{(t-1)} + \mb{p}_o \odot \mb{c}^{(t)} + \mb{b}_o) \\
    \mb{c}^{(t)} & = \mb{o}^{(t)} \odot h(\mb{z}^{(t)}) 
\end{align}
The $i$-th element of the previous vectors in bold corresponds to the value of the gate of the $i$-th neuron of the layer, which is a very convenient and compact representation of the whole layer. Furthermore, if the layer has $N$ LSTM neurons and $M$ inputs (i.e. the size of the layer that precedes this), we see that the input weight matrices $\mb{W}_*$ and the  recurrent weight matrices $\mb{R}_*$ all have size $N \times M$, and that the bias weight matrices $\mb{b}_*$, the peephole weight matrices $\mb{p}_*$ and the matrices $\mb{z}^{(t)}$ through $\mb{c}^{(t)}$ have size $N \times 1$. Although useful for a high-level description in a programming language, the matrix representation  may not be suitable for a direct HDL port. In that case, in order to represent the operation of the $i$-\textit{single neuron}, all the above equations still hold, but instead of \textit{vectors} of outputs, we will have a single scalar output, and we only use the appropriate row in the weight matrices $\mb{W}_{\cdot}$, $\mb{R}_{\cdot}$ and the remaining.

\subsubsection{Training -- SPSA}\label{sec:training_lstm}
Since this work aims to perform on-chip learning, it is important to find a suitable learning scheme. Since I am aiming at an hardware implementation, and although the memory resources of nowadays FPGAs are abundant, one must find an algorithm that uses the less memory as possible (at the smallest performance cost possible), in order to use the additional memory, for instance, to add additional LSTM cells that can improve the performance of our system. 
That being said, we see that the usual training algorithms for LSTM cells -- i.e. Real-Time Recurrent Learning (RTRL), Truncated Backpropagation Through Time (BPTT) or a mixture of both~\cite{Graves05}~\cite{Hoch97}~\cite{Greff15} -- usually involve the storage of the deltas of each layer for every time instant in the training epoch (from $0$ to $T$), which is a highly non-scalable solution both in terms of memory and performance. A most efficient approach to training times series dependant structures like LSTM is the use of \textbf{Simultaneous Perturbation Stochastic Approximation} (SPSA)~\cite{Spall98}. The weight update for the $i$-th weight at the time step $t$ is given by
